Probabilistic inference in dynamical systems

### 2024-04-14 nbs/concepts/inferenceindynamicalsystems/index.qmd
feat(nbs): update inference model without times Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

```{python}
# | label: transcription-splicing-degradation-simulation
import diffrax
from pyrovelocity.models import solve_transcription_splicing_model
from pyrovelocity.models import solve_transcription_splicing_model_analytical
from pyrovelocity.logging import configure_logging
from jax import numpy as jnp

@@

In the case where we can indeed estimate analytical solutions to the system, we
can consider a model with the structure proposed in 
(\ref{eq-init-conds-priors-gene}--\ref{eq-s-obs-cell}) where, given a number of 
genes $G$ and cells $N$, for each
$(i,j) \in \{1, \ldots, G\} \otimes \{1, \ldots, N\}$ we have

@@

    u^{\ast}_{ij} &= 
      1 + (u^{\ast}_{0i} - 1) \cdot e^{-t^{\ast}_j},
      \label{eq-u-star-model-cell} \\

    s^{\ast}_{ij} &= 
      \begin{cases}
        \frac{1}{\gamma^{\ast}_i} + 
        \left( s^{\ast}_{0i} - \xi_i - \frac{1}{\gamma^{\ast}_i} \right) \cdot e^{-\gamma^{\ast}_i t^{\ast}_j} + 
        \xi_i \cdot e^{-t^{\ast}_j},& \gamma^{\ast}_i \neq 1 \\

        1 + (s^{\ast}_{0i} - 1) e^{-t^{\ast}_j} + 
          (u^{\ast}_{0i} - 1) t^{\ast}_j e^{-t^{\ast}_j},& \gamma^{\ast}_i = 1 \\
      \end{cases}, \label{eq-s-star-model-cell} \\

    \hat{u}^{\ast}_{ij} &\sim \text{Normal}(u^{\ast}_{ij}, \sigma_{ui}^2), 
    \label{eq-u-obs-cell} \\

    \hat{s}^{\ast}_{ij} &\sim \text{Normal}(s^{\ast}_{ij}, \sigma_{si}^2). 
    \label{eq-s-obs-cell}
\end{align}

@@

```{python}
# | label: fig-transcription-splicing-degradation-graphical-model-1
# | code-fold: true
# | fig-cap: Graphical model for the transcription-splicing-degradation model.
import daft
import matplotlib.pyplot as mpl

mpl.rcParams["font.family"] = "serif"
mpl.rcParams["font.size"] = 16
mpl.rcParams["text.usetex"] = True

@@

# plates
pgm.add_plate([0.5, 1.2, 5, 4.4], label=r"$i \in \{1, \ldots, G\}$", shift=-0.1)
pgm.add_plate(
    [1.0, 1.8, 5.5, 2.75], label=r"$j \in \{1, \ldots, N\}$", shift=-0.1
)

>>>>

```{python}
# | label: transcription-splicing-degradation-simulation
# | code-fold: true
import diffrax
from pyrovelocity.models import solve_transcription_splicing_model
from pyrovelocity.models import solve_transcription_splicing_model_analytical
from pyrovelocity.logging import configure_logging
from jax import numpy as jnp

@@

In the case where we can indeed estimate analytical solutions to the system, we
can consider a model with the structure proposed in 
(\ref{eq-init-conds-priors-gene}--\ref{eq-s-obs-cell}) where, given a number of 
genes $G$ and cells $N$ that are measured at $K_j$ distinct times, then for each
$(i,j,k) \in \{1, \ldots, G\} \otimes \{1, \ldots, N\} \otimes \{1, \ldots, K_j\}$ 
we have

@@

    {u^{\ast}}^{k}_{ij} &= 
      1 + (u^{\ast}_{0i} - 1) \cdot e^{-{t^{\ast}}^k_j},
      \label{eq-u-star-model-cell} \\

    {s^{\ast}}^{k}_{ij} &= 
      \begin{cases}
        \frac{1}{\gamma^{\ast}_i} + 
        \left( s^{\ast}_{0i} - \xi_i - \frac{1}{\gamma^{\ast}_i} \right) \cdot e^{-\gamma^{\ast}_i {t^{\ast}}^k_j} + 
        \xi_i \cdot e^{-{t^{\ast}}^k_j},& \gamma^{\ast}_i \neq 1 \\

        1 + (s^{\ast}_{0i} - 1) e^{-{t^{\ast}}^k_j} + 
          (u^{\ast}_{0i} - 1) {t^{\ast}}^k_j e^{-{t^{\ast}}^k_j},& \gamma^{\ast}_i = 1 \\
      \end{cases}, \label{eq-s-star-model-cell} \\

    \hat{u}^{\ast}{}^k_{ij} &\sim \text{Normal}({u^{\ast}}^k_{ij}, \sigma_{ui}^2), 
    \label{eq-u-obs-cell} \\

    \hat{s}^{\ast}{}^k_{ij} &\sim \text{Normal}({s^{\ast}}^k_{ij}, \sigma_{si}^2). 
    \label{eq-s-obs-cell}
\end{align}

@@

```{python}
# | label: fig-transcription-splicing-degradation-graphical-model-1
# | code-fold: true
# | fig-cap: Graphical model for the transcription-splicing-degradation model.
import daft
import matplotlib.pyplot as plt

plt.rcParams["font.family"] = "serif"
plt.rcParams["font.size"] = 16
plt.rcParams["text.usetex"] = True

# latent variables for cell-specific outcomes
pgm.add_node(
    "u_star_ij",
    r"${u^{\ast}}^k_{ij}$",
    2,
    3.8,
    scale=1.0,
    shape="rectangle",
)
pgm.add_node(
    "s_star_ij",
    r"${s^{\ast}}^k_{ij}$",
    4,
    3.8,
    scale=1.0,
    shape="rectangle",
)

# observed data
pgm.add_node(
    "t_star_j",
    r"${t^{\ast}}^k_j$",
    5.9,
    3.1,
    observed=True,
    shape="rectangle",
)
pgm.add_node(
    "u_obs_ij",
    r"$\hat{u}^{\ast}{}^{k}_{ij}$",
    2,
    2.4,
    scale=1.0,
    observed=True,
)
pgm.add_node(
    "s_obs_ij",
    r"$\hat{s}^{\ast}{}^{k}_{ij}$",
    4,
    2.4,
    scale=1.0,
    observed=True,
)

# plates
pgm.add_plate(
    [0.4, 1.0, 5, 4.5],
    label=r"$i \in \{1, \ldots, G\}$",
    shift=-0.1,
    fontsize=12,
)
pgm.add_plate(
    [0.8, 1.4, 5.9, 3.2],
    label=r"$j \in \{1, \ldots, N\}$",
    shift=-0.1,
    fontsize=12,
)
pgm.add_plate(
    [1.2, 1.8, 5.2, 2.5],
    label=r"$k \in \{1, \ldots, K_j\}$",
    shift=-0.1,
    fontsize=12,
)

If, for example, we only observe one time point for each cell, 
then @fig-transcription-splicing-degradation-graphical-model-1 and its 
associated description in (\ref{eq-init-conds-priors-gene}--\ref{eq-s-obs-cell}) 
reduces to the graphical model in 
@fig-transcription-splicing-degradation-graphical-model-2.

```{python}
# | label: fig-transcription-splicing-degradation-graphical-model-2
# | code-fold: true
# | fig-cap: Graphical model for the transcription-splicing-degradation model observed at a single time point.
import daft
import matplotlib.pyplot as plt

plt.rcParams["font.family"] = "serif"
plt.rcParams["font.size"] = 16
plt.rcParams["text.usetex"] = True

@@

# plates
pgm.add_plate(
    [0.5, 1.2, 5, 4.4],
    label=r"$i \in \{1, \ldots, G\}$",
    shift=-0.1,
    fontsize=12,
)
pgm.add_plate(
    [1.0, 1.8, 5.5, 2.75],
    label=r"$j \in \{1, \ldots, N\}$",
    shift=-0.1,
    fontsize=12,
)
